---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# clust431

<!-- badges: start -->
<!-- badges: end -->

The goal of clust431 is to ...

## Installation

You can install the released version of clust431 from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("clust431")
```

```{r echo = T, include = F}
library(clust431)
library(tidyverse)
library(mvtnorm)
```

## Example EM Clustering


# **Comparison of EM versus K-means clustering**
```{r}
dataframe = select(iris, Sepal.Length, Sepal.Width)

kmeans(dataframe, 3)
em_clust(dataframe, 3)
```

**These results show that both EM and K-means are producing similar clustering means and clustering vectors; but, EM is also able to provide the covariance matrix for each cluster.**


# **Example of EM clustering in many dimensions**
```{r}
mtcars.cluster = em_clust(mtcars, 4)

mtcars.cluster$Means
mtcars.cluster$Clusters
```


# **Example of EM clustering with a large dataset**
```{r}
dataframe2 = data.frame(x = c(rnorm(500, 20, 5), rnorm(500,1, 10)), 
                        y = c(rnorm(500, 40, 2),rnorm(500, 0, 9)))

em_clust(dataframe2, 2)
ggplot(dataframe2) + geom_point(aes(x, y))
```


# **Example of EM clustering in a single dimension with K-means comparison**
```{r}
em_clust(iris$Sepal.Length, 3)
kmeans(iris$Sepal.Length, 3)
plot(iris$Sepal.Length, pch = 20, ylab = '')
```
